---
title: "NEON forecast challenge submission"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-packages, echo = F, warning=F, message=F}
## install.packages('remotes')
## install.packages('fpp3') # package for applying simple forecasting methods
## install.packages('tsibble') # package for dealing with time series data sets and tsibble objects
## install.packages('tidyverse') # collection of R packages for data manipulation, analysis, and visualisation
## install.packages('lubridate') # working with dates and times
## remotes::install_github('eco4cast/neon4cast') # package from NEON4cast challenge organisers to assist with forecast building and submission

# Load packages`
library(xgboost)
library(tidyverse)
library(lubridate)
library(tidymodels)
```

This is where you could change or add meteorological variables that are used to predict the target

Other variable names can be found at <https://projects.ecoforecast.org/neon4cast-docs/Shared-Forecast-Drivers.html#stage-3>

```{r}

driver_variables <- c("air_temperature")
forecast_variable = "temperature"
```

```{r get-targets, message=F}
#read in the targets data
targets <- read_csv('https://data.ecoforecast.org/neon4cast-targets/aquatics/aquatics-targets.csv.gz')

# read in the sites data
aquatic_sites <- read_csv("https://raw.githubusercontent.com/eco4cast/neon4cast-targets/main/NEON_Field_Site_Metadata_20220412.csv") |>
  dplyr::filter(aquatics == 1)

lake_sites <- aquatic_sites %>%
  filter(field_site_subtype == 'Lake')

# Filter the targets
targets <- targets %>%
  filter(site_id %in% lake_sites$field_site_id,
         variable == forecast_variable)
```

```{r get-NOAA-past, message = F}

# Past stacked weather
noaa_past_s3 <- neon4cast::noaa_stage3()

noaa_past <- noaa_past_s3  |> 
  dplyr::filter(site_id %in% lake_sites$field_site_id,
                datetime >= ymd('2017-01-01'),
                variable %in% driver_variables) |> 
  dplyr::collect()

# aggregate the past to mean values
noaa_past_mean <- noaa_past |>
  mutate(datetime = as_date(datetime)) |>
  group_by(datetime, site_id, parameter, variable) |>
  summarize(prediction = mean(prediction), .groups = "drop") |>
  mutate(prediction = ifelse(variable == driver_variables, prediction - 273.15, prediction)) |>
  pivot_wider(names_from = variable, values_from = prediction) |>
  select(any_of(c('datetime', 'site_id', driver_variables, 'parameter')))
```

```{r get-NOAA-future, message = F}
# Future weather
# New forecast only available at 5am UTC the next day
forecast_date <- Sys.Date() 
noaa_date <- forecast_date - days(1)

noaa_future_s3 <- neon4cast::noaa_stage2(start_date = as.character(noaa_date))

noaa_future <- noaa_future_s3 |> 
  dplyr::filter(datetime >= forecast_date,
                site_id %in% lake_sites$field_site_id,
                variable %in% driver_variables) |> 
  collect()

noaa_future_mean <- noaa_future |>
  mutate(datetime = as_date(datetime)) |>
  group_by(datetime, site_id, parameter, variable) |>
  summarize(prediction = mean(prediction), .groups = "drop") |>
  mutate(prediction = ifelse(variable == "air_temperature", prediction - 273.15, prediction)) |>
  pivot_wider(names_from = variable, values_from = prediction) |>
  select(any_of(c('datetime', 'site_id', driver_variables, 'parameter')))
```

```{r model-setup, message = FALSE}
# Generate a dataframe to fit the model from 
targets_lm <- targets |> 
  pivot_wider(names_from = 'variable', values_from = 'observation') |> 
  left_join(noaa_past_mean, 
            by = c("datetime","site_id"))
```

```{r forecast-loops}
forecast_horizon <- 30
n_members <- 31

lag_days = 1

tidymodels_forecast <- data.frame()

# ----- MAIN LOOP -----

for(i in 1:length(lake_sites$field_site_id)) {  
  # ----- GENERATE AND SPLIT DATASETS -----
  
  site = lake_sites$field_site_id[i]
  #site = "TOOK"  

  # gathers the last day of an observation for each site and how many days out
  # the forecast horizon must go out to
  forecast_start_by_site <- targets |>
      mutate(temperature_lag = lag(observation, lag_days)) |> # account for lag
      dplyr::filter(!is.na(observation) & site_id == site & variable == forecast_variable) |>
      # Start the day after the most recent non-NA value
      dplyr::summarise(start_date = max(datetime)) |>  # Date
      dplyr::mutate(h = (Sys.Date() - start_date) + forecast_horizon,
                    h = as.numeric(h)) |>  # Horizon value
      dplyr::ungroup()

  # this is 1 day after the last date of observation
  forecast_start_date = forecast_start_by_site[[1]]
  # this is the number of days between the start date and today + forecast horizon
  horizon_days = forecast_start_by_site[[2]]
  
  # combine past and future data by site
  noaa_past_site <- noaa_past_mean |>
      dplyr::filter(datetime <= noaa_date,
                  site_id %in% site)
  
  noaa_future_site <- noaa_future_mean |>
      dplyr::filter(datetime > noaa_date,
                    site_id %in% site)
  
  combined_noaa_site <- bind_rows(noaa_past_site, noaa_future_site)

  # generate a data frame to fit the model on
  targets_site <- targets |>
      dplyr::filter(site_id == site,
                    datetime >= min(combined_noaa_site$datetime)) |>
      pivot_wider(names_from = 'variable', values_from = 'observation') |>
      full_join(combined_noaa_site, by = c("datetime","site_id")) |> 
      group_by(parameter) |>
      mutate(temperature_lag = lag(temperature, lag_days)) |>
      ungroup() |>
      mutate(doy = yday(datetime))
  
  # now extract data from 2017-01-01 to date of last observation
  past_to_last_obs <- targets_site |>
    dplyr::filter(datetime < forecast_start_date) |>
    dplyr::filter(!is.na(temperature_lag)) |>
    dplyr::filter(!is.na(temperature))
  
  # extract data from date of last obs + 1 out to forecast horizon
  last_obs_to_horizon <- targets_site |>
    dplyr::filter(datetime >= forecast_start_date)
  
  # ----- Apply Model to Training and Testing Data -----
  
  # split past_to_last_obs for testing and training
  train_data <- past_to_last_obs |>
    dplyr::filter(parameter == 1) |>
    select(-parameter)
    
  #Fit linear model based on past data: water temperature = m * air temperature + b
  #you will need to change the variable on the left side of the ~ if you are forecasting oxygen or chla
  fit <- lm(train_data$temperature ~ train_data$air_temperature + train_data$temperature_lag)
  
  # calculate how well the model fit
  fit_summary <- summary(fit)
  params_se <- fit_summary$coefficients[,2]
  
  mod <- predict(fit, new_data = train_data)
  
  # mod <- c(NA, mod)
  r2 <- round(fit_summary$r.squared, 2) 
  residuals <- mod - train_data$temperature
  err <- mean(residuals, na.rm = TRUE) 
  rmse <- round(sqrt(mean((residuals)^2, na.rm = TRUE)), 2) 
  sigma <- sd(residuals, na.rm = TRUE)
  param_df <- data.frame(beta1 = rnorm(n_members, fit$coefficient[1], params_se[1]),
                       beta2 = rnorm(n_members, fit$coefficient[2], params_se[2]),
                       beta3 = rnorm(n_members, fit$coefficient[3], params_se[3]))

  
  
  targets_future <- last_obs_to_horizon |>
  mutate(prediction = NA,
          doy = yday(datetime),
          parameter = parameter + 1,
          datetime = as_datetime(datetime))
  
  
  prev_day_temps = NA
  # loop through each day
  for (curr_date in unique(targets_future$datetime))
  {
    targets_by_day <- targets_future |>
      dplyr::filter(datetime == curr_date)
    
    first_entry = (curr_date == min(targets_future$datetime))

    if (!first_entry)
    {
      targets_by_day$temperature_lag <- prev_day_temps
    }
    
    targets_by_day$prediction <- param_df$beta1 + 
        param_df$beta2 * targets_by_day$air_temperature + 
        param_df$beta3 * targets_by_day$temperature_lag + 
        rnorm(n = n_members, mean = 0, sd = sigma)
    
    tidymodels_forecast <- bind_rows(tidymodels_forecast, targets_by_day)
    
    prev_day_temps <- targets_by_day$prediction
    
  }
  message(site, ' forecast run')
}
```

Remember to change the model_id when you make changes to the model structure!

```{r}
my_project_id <- "neon4cast"
my_model_id <- 'mkricheldorf_w_uncertainty'

```

```{r make-standard}
# Make forecast fit the EFI standards
tidymodels_forecasts_EFI <- tidymodels_forecast %>%
  mutate(variable = "temperature") |> 
  # For the EFI challenge we only want the forecast for future
  filter(datetime > Sys.Date()) %>%
  group_by(site_id, variable) %>%
  mutate(reference_datetime = min(datetime) - lubridate::days(1),
         family = "ensemble",
         model_id = my_model_id,
         reference_datetime = forecast_date,
         project_id = 'neon4cast',
         duration = 'P1D') %>%
  select(datetime, reference_datetime, site_id, family, parameter, variable, prediction, model_id, project_id, duration)

```

```{r write-forecast}
# Write the forecast to file
theme <- 'aquatics'
forecast_name_1 <- paste0(tidymodels_forecasts_EFI$model_id[1], ".csv")
forecast_file_1 <- paste(theme, forecast_date, forecast_name_1, sep = '-')
forecast_file_1


if (!dir.exists('Forecasts')) {
  dir.create('Forecasts')
}

write_csv(tidymodels_forecasts_EFI, file.path('Forecasts',forecast_file_1))
```

Check that forecast format is valid

```{r}
neon4cast::forecast_output_validator(file.path('Forecasts',forecast_file_1))
```

Change eval = TRUE if you want to submit

```{r submit-forecast, eval= TRUE}

neon4cast::submit(forecast_file = file.path('Forecasts', forecast_file_1), ask = FALSE) # if ask = T (default), it will produce a pop-up box asking if you want to submit

```

```{r plot-forecast}
tidymodels_forecasts_EFI |> 
  ggplot(aes(x=datetime, y=prediction, group = parameter)) +
  geom_line() +
  facet_wrap(~site_id) +
  labs(title = paste0('Forecast generated for ', tidymodels_forecasts_EFI$variable[1], ' on ', tidymodels_forecasts_EFI$reference_datetime[1]))
```

This model applies a simple linear auto-regression model with the equation:
  water_temp_today = forecasted_air_temp_today + forecasted/observed_water_temp_yesterday
  
The driver data used for this is NOAA past 1-day ahead air temperature data and NOAA future N-day ahead air temperature data.

The model was fit using historical temperature observations and historical air temperature 1-day-ahead ensembles from 
the first date which contained water temperature and air temperature data out to the last date an observation was made.

It forecasts temperature from the first date there is no observation outwards to the forecast horizon. It iteratively forecasts 
temperature using the previous day's water temperature and depending on if the day is in the future or past, the model uses
the respective NOAA past mean ensemble data or NOAA future ensemble data.

Although I'm unsure if initial condition uncertainty is quantified correctly, driver uncertainty, parameter uncertainty, and
process uncertainty have been accounted for. The initial fit of the model describes each parameter's standard error and the
standard deviation of the entire model. The parameter standard errors are used to generate a random normal distribution, which are values
used as coefficients for each ensemble. Process uncertainty is accounted for using the standard deviation of the model. An additional
coefficient for noise is added to the end of the modelling equation using this standard deviation. The driver uncertainty is accounted
for in the past and future NOAA ensemble data.

